The Evolution of Phenotypically Invariant Genetic Networks

### 2017-11-14 kalman_walk_paper.tex
note on dimensions

For a more concrete prediction, suppose that the distribution among hybrids is Gaussian.
A population whose trait distribution is Gaussian with mean $\mu$ and variance $\sigma$,
has mean fitness 
\begin{align} \label{eqn:gauss_fit}
\int_{-\infty}^\infty \frac{1}{\sigma\sqrt{2\pi}} e^{-\frac{(x-\mu)^2}{2\sigma^2}} e^{-\frac{x^2}{2\gamma^2}} dx
&=
\sqrt{\frac{1}{1 + \sigma^2/\gamma^2}} \exp\left\{-\frac{\mu^2}{\gamma^2}\left(\frac{1}{1 + \sigma^2/\gamma^2}\right)\right\} .
% &\approx
% \frac{1}{\sigma/\gamma}\left(1 - \frac{\mu^2}{\gamma^2}\right) .
\end{align}
This assumes a single trait, for simplicity; the multivariate case is done in appendix \ref{apx:gauss_load}.
A population of $F_2$s will have, as above, variance $\sigma^2 = \sigma^2_S + 4 \omega \sigma^2_N T/N_e$.
The mean diverges with the square of the distance between the parentals
with a speed depending on the local geometry of the optimal set
(calculated in appendix \ref{apx:H_calc}), so we set $\mu = c_\mu \gamma T/N_e$.
The mean fitness in parental populations is as in equation \ref{eqn:gauss_fit} with $\mu=0$ and $\sigma = \sigma_S$.
This implies that if we define $\fit_2(T)$ to be the mean relative fitness among $F_2$ hybrids
between two populations separated by $T$ generations, 
(i.e., the mean fitness divided by the mean fitness of the parents)
then
\plr{figure out how dimensions come in here}
\begin{align} \label{eqn:fit2}
\fit_2(T) = 
  \left(1 + \frac{4 \omega (\sigma_N/\gamma)^2}{(1 + (\sigma_S/\gamma)^2)} \frac{T}{N_e} \right)^{-1/2} 
     \exp\left\{-\left(c_\mu \frac{T}{N_e}\right)^2 \left(\frac{1}{1 + (\sigma_S/\gamma)^2 + 4 \omega (\sigma_N/\gamma)^2 T/N_e}\right)\right\} .
\end{align}
We discuss the implications of this expression in the next section.

%%%%%%%%%%
\paragraph{Speciation rates under neutrality}
Above, equation \eqref{eqn:fit2} shows how fast hybrids become inviable 
as the time that the parental populations are isolated increases;
what does this tell us about speciation rates under neutrality?
From equation \eqref{eqn:fit2} we can observe that
time is always scaled in units of $N_e$ generations,
the population standard deviations are always scaled by $\gamma$,
and the most important term is the rate of accumulation of segregation variance,
$4 \omega (\sigma_N/\gamma)^2$.
In particular, $\omega$, the rate that hybrid variance increases with separation between parentals,
is proportional to $p-q$, the number of ways mutation can move the system without changing the phenotype,
i.e., the dimensionality of the optimal set.
This implies that less constrained systems should drift faster,
and so lead to speciation more quickly, given the same strength of selection ($\gamma$).
\plr{revisit}
All else being equal, this process will lead to speciation more quickly in smaller populations
and in populations with more neutral genetic variation (larger $\sigma_N$).
These parameters are related -- larger populations generally have more genetic variation --
but since these details depend on the situation, we leave these separate.

@@

Suppose that a population has a Gaussian distribution in trait space
with mean $\mu$ and covariance matrix $\Sigma$,
and that fitness of an individual at $x$ is $\exp(-\|Lx\|^2/2)$.
Then, completing the square as above with $A=\Sigma^{-1}$, $y=\mu$, and $B=L^T L$,
and defining $Q = (\Sigma^{-1} + L^T L)^{-1}$,
\begin{align*}
    &
    \frac{1}{\sqrt{2 \pi}^n \det(\Sigma)^{1/2}} 
        \int e^{-\frac{1}{2} x^T \Sigma^{-1} x} e^{-\frac{1}{2} x^T L^T L x} dx \\
    &\qquad =
    \frac{1}{\sqrt{2 \pi}^n \det(\Sigma)^{1/2}} 
        \int e^{-\frac{1}{2}(x-Q \Sigma^{-1} \mu)^T Q^{-1} (x-Q \Sigma^{-1} \mu)} dx 
        \\ &\qquad \qquad {}
        \times e^{\mu^T \left( I - \Sigma^{-1} Q \right) \Sigma^{-1} \mu} \\
    &\qquad =
    \sqrt{\frac{\det(Q)}{\det(\Sigma)}}
        \exp\left\{\mu^T \left( I - \Sigma^{-1} Q \right) \Sigma^{-1} \mu\right\} \\
    &\qquad =
    \sqrt{\frac{1}{\det(\Sigma)\det(\Sigma^{-1}+L^TL)}}
        \exp\left\{\mu^T \left( I - (I + L^T L \Sigma)^{-1} \right) \Sigma^{-1} \mu\right\} \\
\end{align*}

>>>>

For a more concrete prediction, suppose that the distribution among hybrids is Gaussian.
A population whose trait distribution is Gaussian with mean $\mu$ and variance $\sigma$,
has mean fitness 
\begin{align} \label{eqn:gauss_fit}
\int_{-\infty}^\infty \frac{1}{\sigma\sqrt{2\pi}} e^{-\frac{(x-\mu)^2}{2\sigma^2}} e^{-\frac{x^2}{2\gamma^2}} dx
&=
\sqrt{\frac{1}{1 + \sigma^2/\gamma^2}} \exp\left\{-\frac{\mu^2}{\gamma^2}\left(\frac{1}{1 + \sigma^2/\gamma^2}\right)\right\} .
% &\approx
% \frac{1}{\sigma/\gamma}\left(1 - \frac{\mu^2}{\gamma^2}\right) .
\end{align}
This assumes a single trait, for simplicity; the multivariate case is done in appendix \ref{apx:gauss_load}.
A population of $F_2$s will have, as above, variance $\sigma^2 = \sigma^2_S + 4 \omega \sigma^2_N T/N_e$.
The mean diverges with the square of the distance between the parentals
with a speed depending on the local geometry of the optimal set
(calculated in appendix \ref{apx:H_calc}), so we set $\mu = c_\mu \gamma T/N_e$.
The mean fitness in parental populations is as in equation \ref{eqn:gauss_fit} with $\mu=0$ and $\sigma = \sigma_S$.
This implies that if we define $\fit_2(T)$ to be the mean relative fitness among $F_2$ hybrids
between two populations separated by $T$ generations, 
(i.e., the mean fitness divided by the mean fitness of the parents)
then
\plr{figure out how dimensions come in here}
\begin{align} \label{eqn:fit2}
\fit_2(T) = 
  \left(1 + \frac{4 \omega (\sigma_N/\gamma)^2}{(1 + (\sigma_S/\gamma)^2)} \frac{T}{N_e} \right)^{-1/2} 
     \exp\left\{-\left(c_\mu \frac{T}{N_e}\right)^2 \left(\frac{1}{1 + (\sigma_S/\gamma)^2 + 4 \omega (\sigma_N/\gamma)^2 T/N_e}\right)\right\} .
\end{align}
If each of the $q$ selected directions acts independently,
the drop in fitness will be $\fit_2(T)^q$
(the expression for the correlated cases is given in Appendix \ref{apx:gauss_load}).
We discuss the implications of this expression in the next section.

%%%%%%%%%%
\paragraph{Speciation rates under neutrality}
Above, equation \eqref{eqn:fit2} shows how fast hybrids become inviable 
as the time that the parental populations are isolated increases;
what does this tell us about speciation rates under neutrality?
From equation \eqref{eqn:fit2} we can observe that
time is always scaled in units of $N_e$ generations,
the population standard deviations are always scaled by $\gamma$,
and the most important term is the rate of accumulation of segregation variance,
$4 \omega (\sigma_N/\gamma)^2$.
All else being equal, this process will lead to speciation more quickly in smaller populations
and in populations with more neutral genetic variation (larger $\sigma_N$).
These parameters are related -- larger populations generally have more genetic variation --
but since these details depend on the situation, we leave these separate.

How does this prediction depend on the system size and constraint?
If there are $p$ trait dimensions, constrained in $q$ dimensions,
and if $\omega$ is proportional to $p-q$,
% the rate that hybrid variance increases with separation between parentals, $\omega$,
% is in some cases proportional to $p-q$, 
then the rate that $F_2$ fitness drops is, roughly,
$(1 + 4 (p-q) C T/n_e)^{-q/2} \propto q (p-q)$, where $C$ is a constant.
This suggests that both degree of constraint and number of available neutral directions
affect the speed of accumulation of incompatibilities.
However, note that in real systems, it is likely that $\gamma$ also depends on $p$ and $q$.
\plr{revisit with sims}

@@

Suppose that a population has a Gaussian distribution in $d$-dimensional trait space
with mean $\mu$ and covariance matrix $\Sigma$,
and that fitness of an individual at $x$ is $\exp(-\|Lx\|^2/2)$.
Then, completing the square as above with $A=\Sigma^{-1}$, $y=\mu$, and $B=L^T L$,
and defining $Q = (\Sigma^{-1} + L^T L)^{-1}$,
\begin{align*}
    &
    \frac{1}{\sqrt{2 \pi}^n \det(\Sigma)^{1/2}} 
        \int e^{-\frac{1}{2} x^T \Sigma^{-1} x} e^{-\frac{1}{2} x^T L^T L x} dx \\
    &\qquad =
    \frac{1}{\sqrt{2 \pi}^n \det(\Sigma)^{1/2}} 
        \int e^{-\frac{1}{2}(x-Q \Sigma^{-1} \mu)^T Q^{-1} (x-Q \Sigma^{-1} \mu)} dx 
        \\ &\qquad \qquad {}
        \times e^{\mu^T \left( I - \Sigma^{-1} Q \right) \Sigma^{-1} \mu} \\
    &\qquad =
    \sqrt{\frac{\det(Q)}{\det(\Sigma)}}
        \exp\left\{\mu^T \left( I - \Sigma^{-1} Q \right) \Sigma^{-1} \mu\right\} \\
    &\qquad =
    \sqrt{\frac{1}{\det(\Sigma)\det(\Sigma^{-1}+L^TL)}}
        \exp\left\{\mu^T \left( I - (I + L^T L \Sigma)^{-1} \right) \Sigma^{-1} \mu\right\} \\
\end{align*}

Now suppose that $\Sigma = \sigma^2 I$ and $L = I/\gamma$.
Then,
\begin{align*}
    \sqrt{\frac{1}{\det(\Sigma)\det(\Sigma^{-1}+L^TL)}}
    &=
    \sqrt{\frac{1}{\sigma^{2d} (1/\sigma^2 + 1/\gamma^2)^d}} \\
    &=
    \frac{1}{(1+ \left(\sigma/\gamma\right)^2)^{d/2}} .
\end{align*}
Also, 
\begin{align*}
        \left( I - (I + L^T L \Sigma)^{-1} \right) \Sigma^{-1} 
        &=
        \frac{1}{\sigma^2}\left( 1 - (1 + (\sigma/\gamma)^2)^{-1} \right) I \\
        &=
        \frac{1}{\gamma^2}\frac{1}{(1 + (\sigma/\gamma)^2)} I \\
\end{align*}
